Rising from the Ashes: A Metallicity-Dependent Star Formation Gap Splits the Milky Way's $\alpha$-Sequences

### 2025-02-14 paper/ms.tex
Lars suggestions

\begin{abstract}
    The elemental abundance distribution of stars encodes the history of the gas-phase abundance in the Milky Way. Without a large, unbiased sample of highly precise stellar ages, the exact timing and nature of this history must be \textit{inferred} from the abundances. In the two-dimensional plane of \alphaFe{}-\FeH{}, it is now clear that two separate populations exist -- the low-$\alpha$ and high-$\alpha$ sequences. \edit1{We propose that a brief ($\sim300\,\Myr$) halt in star formation within a narrow metallicity bin can lead to a bimodal \alphaFe{} distribution at that metallicity, assuming a rapidly declining gas phase \alphaFe{}. Using simulations of an idealized setup of a high-$z$ galaxy merger, we show that the merger with the Gaia-Sausage-Enceladus satellite at $z\sim2$ is one possible way to trigger such a gap in the Milky Way. Our proposal does not rely on a merger to induce this age gap. We predict a $\sim300\,\Myr$ gap in stellar ages at a fixed \FeH{} where the $\alpha$-bimodality is prominent ($\FeH\lesssim-0.2$).}
\end{abstract}

@@

Some of the simulations in this orbital grid resulted in bimodal abundance distributions, while some had little to no structure in the abundance distribution plane. \edit1{We will first study two representative simulations in detail chosen based on their structure in the abundance plane as shown in Figure~\ref{fig:fig1}, one which we refer to as bimodal and one as unimodal. For the bimodal simulation, we chose the simulation with $R_0=129\,\kpc$, $V_0=142\,\kms$, and $\eta=0.4$. For the unimodal simulation, the parameters are the same except that $V_0=116\,\kms$. These simulations will later be identified as having the highest and second lowest bimodality score $\mathcal{B}$. We will then examine the full simulation suite, and show the detailed abundance plane for the full suite5 in Appendix~\ref{app:allmerge}.}

@@

\subsection{Direct Search for Quiescence}\label{ssec:obsqui}
Figure~\ref{fig:before_after_sfh_by_iron} indicates a very direct observational test of the mechanism proposed in this work: for disk stars at a given \FeH{}, there should be a gap of $\sim300\,\Myr$ in ages at $\sim8\,\Gyr$, though in the Milky Way the gap could be larger. With a survey of properly chosen stars, this gap could be directly measured with a modestly sized (few hundred) sample of old stars with age uncertainties of \edit1{a few percent}. To our knowledge, the best method at these ages is differential analysis of solar twins, which can provide an age uncertainty of $\sim5\%$ \citep[e.g.][]{2014ApJ...795...23B,2018MNRAS.474.2580S}. However, this has only been applied to stars with solar metallicity, where there is not a clear separation between the high- and low-$\alpha$ sequences (though a gap in ages may still be present). A sort of differential approach could be applied also at lower metallicities, which would simply lack the absolute age calibration that the Sun provides.

@@

\edit1{In this work, we have shown that a brief ($\sim300\,\Myr$) period of halted star formation in a narrow \FeH{} bin is capable of producing a bimodal distribution in \alphaFe{} at that \FeH{}. This proposal requires that the \alphaFe{} of gas within the galaxy is decreasing with time so that the gap in star formation translates to a gap in \alphaFe{}. A global quiescent period can satisfy these constraints, but is not necessary. We demonstrate the plausibility of this scenario using a grid of idealized merger simulations with slightly varied orbital parameters -- however, it is also not necessary that a merger triggers the \FeH{}-dependent halt in star formation.}

@@

\begin{acknowledgements}
We would like to thank Megan Bedell, Christopher Carr, Vedant Chandra, Charlie Conroy, Drummond B. Fielding, Lars Hernquist, Federico Marinacci, Rohan P. Naidu, Melissa K. Ness, Minjung Park, Vadim A. Semenov, and Turner Woody for helpful discussions. We would like to thank Marc Pinsonneault for sharing early access to the APOKASC-3 dataset. We would like to thank Filippo Barbani for kindly providing his version of \texttt{MakeNewDisk}. A.B. would like to thank Todd Phillips for helpful discussions.

>>>>

\begin{abstract}
    The elemental abundance distribution of stars encodes the history of the gas-phase abundance in the Milky Way. Without a large, unbiased sample of highly precise stellar ages, the exact timing and nature of this history must be \textit{inferred} from the abundances. In the two-dimensional plane of \alphaFe{}-\FeH{}, it is now clear that two separate populations exist -- the low-$\alpha$ and high-$\alpha$ sequences. \edit1{We propose that a brief ($\sim300\,\Myr$) halt in star formation within a narrow metallicity bin can lead to a bimodal \alphaFe{} distribution at that metallicity, assuming a rapidly declining gas phase \alphaFe{}. Using simulations of an idealized setup of a high-$z$ galaxy merger, we show that the merger with the Gaia-Sausage-Enceladus satellite at $z\sim2$ is one possible way to trigger such a gap in the Milky Way. This mechanism may also operate in non-merger scenarios. We predict a $\sim300\,\Myr$ gap in stellar ages at a fixed \FeH{} where the $\alpha$-bimodality is prominent ($\FeH\lesssim-0.2$).}
\end{abstract}

@@

Some of the simulations in this orbital grid resulted in bimodal abundance distributions, while some had little to no structure in the abundance distribution plane. \edit1{We will first study two representative simulations in detail chosen based on their structure in the abundance plane as shown in Figure~\ref{fig:fig1}, one which we refer to as bimodal and one as unimodal. For the bimodal simulation, we chose the simulation with $R_0=129\,\kpc$, $V_0=142\,\kms$, and $\eta=0.4$. For the unimodal simulation, the parameters are the same except that $V_0=116\,\kms$. These simulations will later be identified as having the highest and second lowest bimodality score $\mathcal{B}$. We will then examine the full simulation suite, and show the detailed abundance plane for the full suite in Appendix~\ref{app:allmerge}.}

@@

\subsection{Direct Observational Test}\label{ssec:obsqui}
Figure~\ref{fig:before_after_sfh_by_iron} indicates a very direct observational test of the mechanism proposed in this work: for disk stars at a given \FeH{}, there should be a gap of $\sim300\,\Myr$ in ages at $\sim8\,\Gyr$, though in the Milky Way the gap could be larger. With a survey of properly chosen stars, this gap could be directly measured with a modestly sized (few hundred) sample of old stars with age uncertainties of \edit1{a few percent}. To our knowledge, the best method at these ages is differential analysis of solar twins, which can provide an age uncertainty of $\sim5\%$ \citep[e.g.][]{2014ApJ...795...23B,2018MNRAS.474.2580S}. However, this has only been applied to stars with solar metallicity, where there is not a clear separation between the high- and low-$\alpha$ sequences (though a gap in ages may still be present). A sort of differential approach could be applied also at lower metallicities, which would simply lack the absolute age calibration that the Sun provides.

@@

\edit1{In this work, we have shown that a brief ($\sim300\,\Myr$) period of halted star formation in a narrow \FeH{} bin is capable of producing a bimodal distribution in \alphaFe{} at that \FeH{}. This proposal requires that the \alphaFe{} of gas within the galaxy is decreasing with time so that the gap in star formation translates to a gap in \alphaFe{}. A global quiescent period can satisfy these constraints, but is not necessary. We demonstrate the plausibility of this scenario using a grid of idealized merger simulations with slightly varied orbital parameters. This scenario could potentially be triggered in non-merger scenarios, which we plan to explore in future work.}

@@

\begin{acknowledgements}
\edit1{We would like to thank the anonymous referee for their helpful comments which have served to clarify and strengthen our main arguments.} We would like to thank Megan Bedell, Christopher Carr, Vedant Chandra, Charlie Conroy, Drummond B. Fielding, Lars Hernquist, Federico Marinacci, Rohan P. Naidu, Melissa K. Ness, Minjung Park, Vadim A. Semenov, and Turner Woody for helpful discussions. We would like to thank Marc Pinsonneault for sharing early access to the APOKASC-3 dataset. We would like to thank Filippo Barbani for kindly providing his version of \texttt{MakeNewDisk}. A.B. would like to thank Todd Phillips for helpful discussions.
